---
title: "Classifying Text with Local Models (Ollama)"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, comment = "#>", collapse = TRUE)
```

# Classify text with a local Ollama model

When you don't want to pay for cloud API calls or your data can't leave your machine (IRB, HIPAA, internal policy), run classification entirely locally via [Ollama](https://ollama.com).

## 1. Install Ollama and pull a model

In a terminal:

```bash
# Install: see https://ollama.com/download
ollama pull qwen2.5:7b
ollama serve   # runs in the background; one-time
```

## 2. Install the R package

```{r install}
install.packages("cat.stack",
                 repos = c("https://chrissoria.r-universe.dev",
                          "https://cloud.r-project.org"))
library(cat.stack)
```

## 3. Classify with the local model

No `api_key` is needed when using Ollama:

```{r classify}
responses <- c(
  "The new wellness program is great, I've been using it daily.",
  "It's confusing and the app crashes constantly.",
  "I haven't tried it yet but it sounds promising."
)

results <- classify(
  input_data    = responses,
  categories    = c("Positive", "Negative", "Neutral"),
  user_model    = "qwen2.5:7b",
  model_source  = "ollama"
)

print(results)
```

## 4. Why local?

- **Zero API cost** — only your machine's electricity.
- **Data never leaves your network** — important for sensitive survey data, clinical text, internal documents.
- **Reproducible** — model weights are pinned by `ollama pull <model>:<tag>` and don't change underneath you the way cloud models do.

The trade-off: local 7B–14B models are usually a few percentage points behind frontier cloud models on classification accuracy. For most survey work the gap is small; for high-stakes coding, validate against a labeled subsample.

## 5. Suggested local models

| Model               | Pull command                | Notes                                  |
|---------------------|-----------------------------|----------------------------------------|
| `qwen2.5:7b`        | `ollama pull qwen2.5:7b`    | Good baseline; fast on Apple Silicon   |
| `qwen2.5:14b`       | `ollama pull qwen2.5:14b`   | Higher accuracy; ~10 GB RAM            |
| `llama3.1:8b`       | `ollama pull llama3.1:8b`   | Solid generalist                       |
| `gpt-oss:20b`       | `ollama pull gpt-oss:20b`   | Strong for English, slower             |

See https://ollama.com/library for the full catalog.
