---
title: "Ensemble Classification with Cloud and Local Models"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, comment = "#>", collapse = TRUE)
```

# Multi-model ensemble classification

Running the same input through multiple models and combining via majority voting often improves accuracy by reducing individual model bias. Per the CatLLM evaluation, inexpensive open-weight ensembles can match or exceed individual frontier-class models at a fraction of the cost.

## Set up

```{r setup-pkgs}
library(cat.stack)
# Have OPENAI_API_KEY, ANTHROPIC_API_KEY, GOOGLE_API_KEY in .Renviron
```

## Cloud-only ensemble (3 providers, unanimous voting)

```{r cloud-ensemble}
responses <- c(
  "Took a new job in Chicago",
  "Wanted to be closer to grandkids",
  "Couldn't afford rent in the Bay Area",
  "Job market collapsed after the layoffs",
  "Family pressure to move home"
)

verbose_cats <- c(
  "Job/school: A change in employment, education, or career, including transfers and retirement.",
  "Family: Relationship changes, having children, supporting relatives, or relocating to be near family.",
  "Cost of living: Housing affordability, cost of goods, or general economic pressure.",
  "Other: The response does not fit any of the above categories."
)

results <- classify(
  input_data = responses,
  categories = verbose_cats,
  models = list(
    c("gpt-4o-mini",              "openai",    Sys.getenv("OPENAI_API_KEY")),
    c("claude-3-5-haiku-20241022", "anthropic", Sys.getenv("ANTHROPIC_API_KEY")),
    c("gemini-2.5-flash",         "google",    Sys.getenv("GOOGLE_API_KEY"))
  ),
  consensus_threshold = "unanimous"   # alternatives: "majority", 0.5, etc.
)

# The output includes per-model columns + a consensus column
str(results)
```

## Cloud + local hybrid ensemble

Mixing cloud and local models gives you the speed of cloud + the cost/privacy benefits of local. Each model's vote counts equally.

```{r hybrid-ensemble}
results <- classify(
  input_data = responses,
  categories = verbose_cats,
  models = list(
    c("gpt-4o-mini",              "openai",    Sys.getenv("OPENAI_API_KEY")),
    c("claude-3-5-haiku-20241022", "anthropic", Sys.getenv("ANTHROPIC_API_KEY")),
    c("qwen2.5:7b",               "ollama",    "")           # local, no key
  ),
  consensus_threshold = "majority"   # 2 of 3 agree
)
```

## Consensus thresholds

| Threshold           | Meaning                                                                  |
|---------------------|--------------------------------------------------------------------------|
| `"unanimous"`       | Only assign category if every model agrees. Highest precision (default).  |
| `"majority"`        | Assign if more than half agree (50%). Good balance.                       |
| `"two-thirds"`      | At least two-thirds must agree (~67%).                                     |
| Custom numeric `0.0`–`1.0` | Any threshold between 0 and 1.                                       |

## Output columns

For an ensemble with N models classifying into K categories, the result `data.frame` includes:

- `input_data` — original text
- `category_<i>_<model_slug>` columns (i = 1..K) for each (model × category) pair
- `category_<i>_consensus` columns with the voted result (0/1)
- Agreement scores showing how many models agreed

Use the `_consensus` columns for downstream analysis; the per-model columns help diagnose disagreement.
