---
title: "Validating a Hybrid Cloud + Local Ensemble"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, comment = "#>", collapse = TRUE)
```

# Validate the ensemble pipeline end-to-end

This walkthrough exercises the multi-model ensemble path on a real
classification task and verifies the output `data.frame` has the
expected structure. Useful as:

- A **smoke test** after install / upgrade — does the ensemble flow
  still work?
- A **proof of behavior** before betting a real research run on it.

It also exercises the **Ollama auto-start** wiring and the **JSON
formatter auto-enable** that fires whenever an ensemble entry uses the
`"ollama"` provider.

## Prereqs

```{r prereqs}
# Install the ecosystem
install.packages("cat.llm",
                 repos = c("https://chrissoria.r-universe.dev",
                          "https://cloud.r-project.org"))
library(cat.llm)

# Python backend (one-time)
install_cat_stack()

# Local models for the test (one-time pulls in a terminal):
#   ollama pull qwen2.5:7b      # ~4.7 GB
#   ollama pull mistral:7b      # ~4.1 GB
#   ollama pull phi3:mini       # ~2.3 GB
```

You do **not** need to start Ollama manually — `classify()` auto-starts
it on first use because at least one ensemble entry has
`"ollama"` as the provider.

## A small hybrid ensemble

A mix of 3 local models (different vendors) + an optional cloud model
keeps the test fast and exercises both code paths.

```{r ensemble}
verbose_cats <- c(
  "Positive: The respondent expresses satisfaction, approval, or favorable sentiment.",
  "Negative: The respondent expresses dissatisfaction, frustration, or criticism.",
  "Neutral: The respondent is factual, ambivalent, or does not express clear sentiment.",
  "Other: The response does not fit any of the above categories."
)

responses <- c(
  "The new wellness program is great, I've been using it daily.",
  "It's confusing and the app crashes constantly.",
  "I haven't tried it yet but it sounds promising."
)

results <- classify(
  input_data = responses,
  categories = verbose_cats,
  models = list(
    c("qwen2.5:7b", "ollama", ""),
    c("mistral:7b", "ollama", ""),
    c("phi3:mini",  "ollama", "")
    # Add a cloud model for a hybrid ensemble:
    # c("gpt-4o-mini", "openai", Sys.getenv("OPENAI_API_KEY"))
  ),
  consensus_threshold = "majority",   # 2 of 3 must agree
  check_verbosity     = FALSE
)
```

Expect the console to show, in order:

```
[CatLLM] Ollama detected — auto-enabling JSON formatter fallback
  (small local models more often emit malformed JSON).
  Pass json_formatter=False to opt out.

Validating 3 model configuration(s)...
Classifying responses: 100% ...
```

If any row's JSON parse fails, you'll additionally see:

```
[CatLLM] First malformed-JSON row encountered — loading
  JSON formatter model into RAM now (one-time per session).
[CatLLM] Loading JSON formatter on cpu...
[CatLLM] JSON formatter ready.
```

If every row parses cleanly, the formatter stays unloaded — zero cost.

## Validate the output structure

Check that the result has the expected columns:

```{r assertions}
expected_cols <- c(
  "input_data",
  "processing_status",
  # Per-model columns: category_<i>_<model_slug>
  "category_1_qwen2_5_7b_tauto",
  "category_1_mistral_7b_tauto",
  "category_1_phi3_mini_tauto",
  # Consensus columns: category_<i>_consensus
  "category_1_consensus",
  "category_2_consensus",
  "category_3_consensus",
  "category_4_consensus",
  # Agreement scores: category_<i>_agreement (fraction of models that agreed)
  "category_1_agreement"
)

stopifnot(all(expected_cols %in% colnames(results)))
stopifnot(nrow(results) == length(responses))
stopifnot(all(results$processing_status %in% c("success", "partial")))

# Consensus columns should be 0/1
consensus_cols <- grep("_consensus$", colnames(results), value = TRUE)
stopifnot(all(unlist(results[consensus_cols]) %in% c(0, 1)))

# Agreement columns should be in [0, 1]
agreement_cols <- grep("_agreement$", colnames(results), value = TRUE)
stopifnot(all(unlist(results[agreement_cols]) >= 0 &
              unlist(results[agreement_cols]) <= 1))

cat("All structural assertions passed.\n")
```

## Inspect the consensus

```{r inspect}
print(results[, c("input_data", "processing_status",
                  grep("_consensus$", colnames(results), value = TRUE))])
```

A reasonable run with all 3 local models will get the first row right
("Positive") and possibly miss row 2 ("Negative") because the smaller
models sometimes return all-zeros on short critical text. That's a
known model-quality issue, not a cat-llm bug — adding a stronger model
(e.g. `gpt-4o-mini`) to the ensemble closes the gap.

## What this confirmed

- Ollama auto-start fires when the server isn't running
- `json_formatter` defaults to `TRUE` when any ensemble entry uses
  Ollama
- Formatter model loads only on first parse failure (lazy)
- The output `data.frame` has the documented per-model + consensus +
  agreement column shape

For deeper conceptual coverage of ensemble mode (consensus thresholds,
hybrid cloud+local, the per-model output columns), see
`03-ensemble-classification-cloud-and-local.Rmd`.
