---
title: "Exploring Categories with explore()"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, comment = "#>", collapse = TRUE)
```

# Use `explore()` for saturation analysis

`explore()` is the raw category-extraction step. Unlike `extract()` (which de-duplicates and merges semantically similar labels), `explore()` returns **every category string from every chunk across every iteration** — with duplicates intact. That makes it the right tool for understanding **category stability**: do the same themes keep appearing, or are new ones still being discovered with each iteration?

## Use case

You have ~500 open-ended survey responses and want to know: "If I sample 50 of them, am I capturing most of the themes? Or do I need to look at all 500?"

## Set up

```{r setup-pkgs}
# install.packages("cat.llm",
#                  repos = c("https://chrissoria.r-universe.dev",
#                            "https://cloud.r-project.org"))
library(cat.llm)
api_key <- Sys.getenv("OPENAI_API_KEY")
```

## Run explore with several iterations

`explore_survey()` is cat.llm's survey-domain alias for
`cat.survey::explore()` — it injects the survey-question framing into
the prompt for more accurate theme discovery.

```{r explore}
raw_categories <- explore_survey(
  input_data           = df$open_response,
  survey_question      = "Why did you move to this city?",
  iterations           = 12L,         # number of passes
  divisions            = 8L,          # data chunks per pass
  categories_per_chunk = 10L,         # max categories pulled per chunk
  api_key              = api_key,
  user_model           = "gpt-4o-mini"
)

length(raw_categories)
head(raw_categories, 20)
```

## Analyze stability

Count how often each raw category string appeared. Themes that show up often are stable; themes that show up once are likely chunk-specific noise.

```{r stability}
cat_counts <- sort(table(raw_categories), decreasing = TRUE)
head(cat_counts, 20)

# Themes appearing in at least 25% of iterations are "stable"
stable_themes <- names(cat_counts[cat_counts >= max(cat_counts) * 0.25])
stable_themes
```

## Saturation plot

If the rate of new unique categories per iteration is dropping toward zero, you've reached saturation.

```{r saturation, eval = FALSE}
# Hypothetically split your responses into 4 sample sizes and re-run:
sample_sizes <- c(50, 100, 250, 500)
unique_counts <- sapply(sample_sizes, function(n) {
  raw <- explore_survey(
    input_data           = sample(df$open_response, n),
    survey_question      = "Why did you move to this city?",
    iterations           = 4L,
    api_key              = api_key,
    user_model           = "gpt-4o-mini"
  )
  length(unique(raw))
})

plot(sample_sizes, unique_counts, type = "b",
     xlab = "Sample size", ylab = "Unique categories discovered")
```

A flattening curve means you've hit saturation; a still-rising curve means more data is still revealing new themes.

## When to use `explore()` vs `extract()`

| Function    | Output                                  | Best for                                                    |
|-------------|-----------------------------------------|-------------------------------------------------------------|
| `explore()` | Raw category strings (with duplicates)   | Saturation analysis, finding stable themes, methods papers   |
| `extract()` | De-duplicated, normalized top categories | Building a clean codebook to pass into `classify()`         |
