---
title: "Extracting Categories with extract()"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, comment = "#>", collapse = TRUE)
```

# Discover a category scheme with `extract()`

`extract()` discovers and returns a **normalized, de-duplicated** set of categories from your data. Use this when you have free-text data and no pre-defined coding scheme — let the model surface the recurring themes for you.

Behind the scenes, `extract()` chunks your data, runs the discovery prompt multiple times (across `iterations`), merges semantically similar labels, and returns a ranked list of top categories.

## Set up

```{r setup-pkgs}
library(cat.stack)
api_key <- Sys.getenv("OPENAI_API_KEY")
```

## Basic usage

```{r basic-extract}
responses <- c(
  "Just had the best coffee ever! Highly recommend the new place downtown.",
  "The bus was 40 minutes late again. Public transit here is a joke.",
  "Loved the museum exhibit on California water rights.",
  "Why is rent so expensive in this neighborhood?",
  "Tried the new ramen spot — totally worth the wait.",
  "Construction noise on 4th street has been brutal lately.",
  "The library's new author series is amazing.",
  "Need to find an affordable place before lease ends.",
  "Saw a great documentary at the indie theater last night.",
  "Trash pickup hasn't come in three weeks. Calling 311."
)

result <- extract(
  input_data           = responses,
  api_key              = api_key,
  user_model           = "gpt-4o-mini",
  max_categories       = 8L,
  iterations           = 6L
)

result$top_categories
```

## Inspect the full output

`extract()` returns a named list with three components:

```{r inspect}
names(result)
# [1] "counts_df"      "top_categories"  "raw_top_text"

# counts_df: every candidate category + how many iterations picked it
head(result$counts_df, 15)

# top_categories: the curated, ordered list (this is what you'd pass to classify())
result$top_categories

# raw_top_text: the original prose the model produced (useful for debugging prompt issues)
head(result$raw_top_text, 3)
```

## Tune the discovery

| Argument               | Effect                                                                          |
|------------------------|--------------------------------------------------------------------------------|
| `max_categories`       | Cap on the number of categories returned. Smaller = forces the model to merge.   |
| `categories_per_chunk` | Categories the model proposes per chunk before consolidation.                    |
| `divisions`            | Number of chunks the data is split into per iteration.                            |
| `iterations`           | Number of passes. More passes = more stable but more expensive.                   |
| `specificity`          | `"broad"` (default) or `"specific"`. `"specific"` gives finer-grained themes.    |

```{r tuned-extract}
result_specific <- extract(
  input_data           = responses,
  api_key              = api_key,
  user_model           = "gpt-4o-mini",
  max_categories       = 12L,
  iterations           = 10L,
  specificity          = "specific"
)
result_specific$top_categories
```

## What to do with the output

Pass `result$top_categories` directly to `classify()`:

```{r classify-from-extract}
classified <- classify(
  input_data = responses,
  categories = result$top_categories,
  api_key    = api_key,
  user_model = "gpt-4o-mini"
)
```

Or curate manually first — rewriting the labels to be more verbose tends to boost classification accuracy by several percentage points (see the Python README's "Best Practices for Classification" section).
