---
title: "Extracting Categories from Columns of Text"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, comment = "#>", collapse = TRUE)
```

# End-to-end: data.frame column → classified dataset

The canonical CatLLM workflow when you don't have a pre-defined coding scheme:

1. `extract()` to discover the category scheme from your data
2. `classify()` to assign every row to one or more of those categories
3. Export to CSV for analysis in R, Stata, or Python

## Set up

```{r setup-pkgs}
# install.packages("cat.llm",
#                  repos = c("https://chrissoria.r-universe.dev",
#                            "https://cloud.r-project.org"))
library(cat.llm)
library(readr)

api_key <- Sys.getenv("OPENAI_API_KEY")

df <- read_csv("survey_responses.csv")   # has a column `open_response`
head(df$open_response)
```

## Step 1 — Extract a category scheme

`extract_survey()` is cat.llm's survey-domain alias for
`cat.survey::extract()` — survey-question framing is built in.

```{r extract}
cats <- extract_survey(
  input_data           = df$open_response,
  survey_question      = "Why did you move to a new city?",
  max_categories       = 12L,
  api_key              = api_key,
  user_model           = "gpt-4o-mini"
)

cats$top_categories
```

Review `cats$top_categories` and edit if needed — you can drop, merge, or rewrite category labels before classifying. **Verbose labels classify more accurately than short ones.**

```{r curate-categories}
verbose_cats <- c(
  "Job/school: A change in employment, education, or career, including transfers and retirement.",
  "Family: Relationship changes, having children, supporting relatives, or relocating to be near family.",
  "Cost of living: Housing affordability, cost of goods, or general economic pressure.",
  "Lifestyle: Climate, recreation, cultural fit, or change of scenery.",
  "Other: The response does not fit any of the above categories."
)
```

## Step 2 — Classify every row

```{r classify}
results <- classify_survey(
  input_data      = df$open_response,
  categories      = verbose_cats,
  survey_question = "Why did you move to a new city?",
  api_key         = api_key,
  user_model      = "gpt-4o-mini"
)

str(results)
```

The output `data.frame` has one row per input response and one column per category (0/1 indicators by default, with the `_text` column preserving the original input).

## Step 3 — Merge back into the original data.frame

```{r merge}
df_classified <- cbind(df, results)
head(df_classified)
```

## Step 4 — Export and analyze

```{r export}
write_csv(df_classified, "survey_responses_classified.csv")

# Or jump into analysis right away:
library(dplyr)
df_classified %>%
  dplyr::summarize(
    across(all_of(c("Job/school", "Family", "Cost of living", "Lifestyle", "Other")), mean)
  )
```

## Tips

- **Run on a subsample first** to verify the category scheme makes sense before classifying thousands of rows.
- **Use ensemble mode** (see `03-ensemble-classification-cloud-and-local.Rmd`) for high-stakes coding.
- **Validate against a hand-labeled subsample** for any published research — see the Python README's "Best Practices" section for guidance.
