---
title: "Summarizing Text and PDF Data"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, comment = "#>", collapse = TRUE)
```

# Summarize text and PDFs with `summarize()`

`summarize()` produces structured summaries of text, PDFs, or images. Useful for digesting long policy documents, academic abstracts, or a corpus of social media posts into a comparable summary column you can analyze.

> **Note:** Calls below use the namespaced form `cat.stack::summarize()` to avoid collision with `dplyr::summarize()`, `base::summary()`, and other `summarize` aliases. After `library(cat.llm)` only, the bare `summarize()` (re-exported from `cat.stack`) is unambiguous; the explicit form is safer in mixed sessions.

## Set up

```{r setup-pkgs}
# install.packages("cat.llm",
#                  repos = c("https://chrissoria.r-universe.dev",
#                            "https://cloud.r-project.org"))
library(cat.llm)
api_key <- Sys.getenv("OPENAI_API_KEY")
```

## Summarize text

```{r summarize-text}
docs <- c(
  paste("The city council voted 6-3 to approve the new affordable housing ordinance,",
        "which requires developers of buildings with 10+ units to set aside 15% as",
        "affordable to households earning less than 80% of area median income..."),
  paste("After three months of public comment, the planning commission has adopted",
        "new zoning rules for the downtown corridor, easing parking minimums for",
        "buildings within 1,000 feet of a transit stop...")
)

results <- cat.stack::summarize(
  input_data   = docs,
  format       = "paragraph",   # "paragraph", "bullets", "one_liner", "structured", "full"
  instructions = "Focus on what the policy does and who it affects",
  api_key      = api_key,
  user_model   = "gpt-4o-mini"
)

results$summary
```

## Summarize PDFs

For PDF input, install the PDF extras and pass a directory of PDFs:

```{r install-pdf}
# One-time:
# cat.stack::install_cat_stack(pdf = TRUE)
```

```{r summarize-pdf}
results <- cat.stack::summarize(
  input_data   = "./policy_pdfs/",     # directory of PDFs (or vector of file paths)
  format       = "bullets",
  instructions = "Summarize each PDF as 5 bullet points covering: what, who, why, when, cost.",
  api_key      = api_key,
  user_model   = "gpt-4o-mini",
  pdf_dpi      = 150L
)
```

## Output formats

| `format`       | Output style                                                              |
|----------------|---------------------------------------------------------------------------|
| `"paragraph"`  | 2–4 sentence prose summary                                                |
| `"bullets"`    | 3–6 short bullet points                                                   |
| `"one_liner"`  | Single-sentence headline                                                  |
| `"structured"` | JSON-like fields (e.g. `topic`, `audience`, `key_finding`)                |
| `"full"`       | Longer multi-paragraph summary suitable for review                        |

## Multi-model ensemble for summaries

Same `models = list(...)` pattern as `classify()` works here too:

```{r ensemble-summarize}
results <- cat.stack::summarize(
  input_data = docs,
  format     = "bullets",
  models = list(
    c("gpt-4o-mini",              "openai",    Sys.getenv("OPENAI_API_KEY")),
    c("claude-3-5-haiku-20241022", "anthropic", Sys.getenv("ANTHROPIC_API_KEY"))
  )
)
```

For summaries, ensembles produce per-model columns (e.g. `summary_gpt_4o_mini`, `summary_claude`) rather than a voted consensus — summarization isn't a discrete classification, so use the per-model outputs to compare or merge manually.
