---
title: "Extracting and Exploring Categories with Local Models (Ollama)"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, comment = "#>", collapse = TRUE)
```

# Extracting and Exploring Categories with Local Models (Ollama)

This walkthrough runs `extract()` and `explore()` entirely on your local
machine via Ollama — no API keys, no cloud costs, full data privacy.

What this script covers:

1. Checking Ollama status and listing installed models
2. `extract()` — normalized, deduplicated category list
3. `explore()` — raw category frequency list for saturation analysis

## Prerequisites

```{r prereqs}
# install.packages("cat.llm",
#                  repos = c("https://chrissoria.r-universe.dev",
#                            "https://cloud.r-project.org"))
library(cat.llm)

# One-time Python backend install:
# install_cat_stack()

# One-time model pull (in a terminal):
#   ollama pull llama3.2      # ~2.0 GB
#   ollama pull qwen2.5:7b   # ~4.7 GB
```

## 1. Check Ollama

`ensure_ollama_running()` checks whether Ollama is reachable and,
if not, attempts to start it automatically (macOS: via the Ollama.app
daemon; Linux/Windows: via `ollama serve`).

```{r check-ollama}
ensure_ollama_running()

# See what models you have installed
models <- list_ollama_models()
cat("Installed models:", length(models), "\n")
for (m in models) cat(" -", m, "\n")

# Set the model you want to use — change to any model you have installed
MODEL <- "llama3.2"

if (!check_ollama_model(MODEL)) {
  cat("'", MODEL, "' is not installed. Downloading now...\n", sep = "")
  pull_ollama_model(MODEL, auto_confirm = TRUE)
}
```

`check_ollama_model()` does partial matching, so `"llama3.2"` matches
`"llama3.2:latest"`.  `pull_ollama_model()` prints estimated size and a
resource check before downloading; `auto_confirm = TRUE` skips the
interactive prompt — safe in scripts and RMarkdown documents.

## 2. Load data

```{r load-data}
# Using a character vector here; in practice substitute your own CSV column:
# df <- read.csv("survey_responses.csv")
# responses <- df$open_response

responses <- c(
  "I moved to be closer to family.",
  "Better job opportunities in this city.",
  "Lower cost of living.",
  "Wanted a warmer climate.",
  "My partner got into graduate school here.",
  "I was priced out of my old neighborhood.",
  "Looking for a fresh start after my divorce.",
  "Company relocated our office.",
  "Needed more space for a home office.",
  "Better schools for the kids.",
  "I retired and wanted to be near the beach.",
  "Cheaper rent and more outdoor activities.",
  "Transferred to a new branch.",
  "Family illness — needed to be nearby.",
  "Just wanted to live somewhere more walkable."
)

SURVEY_QUESTION <- "Why did you move to your current residence?"
```

## 3. `extract()` — normalized category list

`extract()` runs multiple iterations of chunked extraction, semantically
merges the results, and returns a ranked, deduplicated list. Use this as
the first step before `classify()`.

```{r extract}
extract_results <- extract(
  input_data      = responses,
  survey_question = SURVEY_QUESTION,
  user_model      = MODEL,
  model_source    = "ollama",
  max_categories  = 10L,
  iterations      = 3L
)

cat("Top categories (extract):\n")
for (cat_label in extract_results$top_categories) {
  cat(" ", cat_label, "\n")
}

cat("\nCategory frequency counts:\n")
print(head(extract_results$counts_df, 15))
```

The three components of the return value:

| Field            | Contents                                                       |
|------------------|----------------------------------------------------------------|
| `top_categories` | Curated, ordered character vector — ready to pass to `classify()` |
| `counts_df`      | Every candidate category + how many iterations selected it     |
| `raw_top_text`   | Raw model prose (useful for debugging prompt issues)           |

## 4. `explore()` — raw category list for saturation analysis

`explore()` returns *every* category string from every chunk across every
iteration — duplicates intact. Use this to check whether your category
space is saturating (i.e. whether new iterations stop producing new
categories).

```{r explore}
raw_categories <- explore(
  input_data      = responses,
  survey_question = SURVEY_QUESTION,
  user_model      = MODEL,
  model_source    = "ollama",
  iterations      = 3L
)

cat("Total raw extractions:", length(raw_categories), "\n")
cat("Unique categories:    ", length(unique(raw_categories)), "\n")
```

### Saturation check

New unique categories should decrease with each iteration. If iteration 3
still introduces as many new categories as iteration 1, add more
iterations.

```{r saturation}
n_iter <- 3L
per_iter <- length(raw_categories) %/% n_iter
seen <- character(0)

cat("Saturation by iteration:\n")
for (i in seq_len(n_iter)) {
  chunk <- raw_categories[((i - 1L) * per_iter + 1L):(i * per_iter)]
  new_cats <- setdiff(chunk, seen)
  seen <- union(seen, chunk)
  cat(sprintf("  Iteration %d: %3d new unique categories\n", i, length(new_cats)))
}

cat("\nSample raw categories (first 20):\n")
for (c in head(raw_categories, 20)) cat(" -", c, "\n")
```

## From explore/extract to classify

Pass `extract_results$top_categories` directly to `classify()`, or
rewrite the labels to be more verbose first (verbose labels typically
improve accuracy by several percentage points):

```{r classify}
verbose_cats <- c(
  "Employment/Career: Moving for a new job, promotion, transfer, or better career prospects.",
  "Family/Relationships: Relocating to be near family, following a partner, or after a major life event.",
  "Cost/Affordability: Driven by housing costs, rent increases, or seeking a cheaper area.",
  "Lifestyle/Environment: Seeking a preferred climate, outdoor access, walkability, or quality of life.",
  "Education: Moving for better schools for children, or a partner's graduate program.",
  "Other: The response does not fit any of the above categories."
)

results <- classify(
  input_data      = responses,
  categories      = verbose_cats,
  survey_question = SURVEY_QUESTION,
  user_model      = MODEL,
  model_source    = "ollama"
)

str(results)
```

## Tips

- **Ollama auto-starts automatically** when you call `classify()` or
  `extract()` with `model_source = "ollama"` — you don't need to call
  `ensure_ollama_running()` manually, though doing so first gives a clear
  error if Ollama isn't installed.
- **Small models (phi3:mini, qwen2.5:3b) work** for extraction;
  classification accuracy improves with larger models (qwen2.5:7b,
  mistral:7b).
- **Use ensemble mode** with multiple local models for higher-stakes
  coding — see `02-validate-hybrid-ensemble.Rmd` and
  `03-ensemble-classification-cloud-and-local.Rmd`.
