---
title: "Automatic Prompt Optimization with prompt_tune()"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, comment = "#>", collapse = TRUE)
```

# Automatic Prompt Optimization with `prompt_tune()`

`prompt_tune()` iteratively refines the classification system prompt by
analyzing category-level errors in a small labeled sample. Each iteration:

1. Classifies a random sample with the current prompt
2. Opens a browser window so you can correct misclassifications
3. Asks an LLM to generate targeted per-category instructions from your corrections
4. Re-classifies the sample with the updated prompt
5. Keeps the prompt only if accuracy improves

Categories are **never modified** — only the system prompt changes.

**What this script covers:**

1. Basic `prompt_tune()` with a cloud model
2. Understanding the output (tuned prompt, accuracy before/after)
3. Applying the tuned prompt to a full dataset via `classify()`
4. Tips for getting the most out of prompt tuning

**What you need:**

- `library(cat.llm)` with Python backend installed
- An API key (OpenAI used here; any supported provider works)
- A browser (the correction UI opens automatically)

**Cost:** ~$0.10–0.30 per tuning run (a handful of small model calls)

**Runtime:** ~2–5 minutes including your time labeling corrections

---

## Prerequisites

```{r prereqs}
# install.packages("cat.llm",
#                  repos = c("https://chrissoria.r-universe.dev",
#                            "https://cloud.r-project.org"))
library(cat.llm)

# One-time Python backend install:
# install_cat_stack()

api_key <- Sys.getenv("OPENAI_API_KEY")   # or paste your key directly
```

## Sample data

```{r data}
responses <- c(
  "The new wellness program is great, I've been using it daily.",
  "It's confusing and the app crashes constantly.",
  "I haven't tried it yet but it sounds promising.",
  "The interface is intuitive and I like the dashboard.",
  "Customer support never responded to my emails.",
  "Pretty average -- nothing special either way.",
  "Best benefit my employer has ever offered.",
  "I tried it once and forgot about it.",
  "The login keeps failing on my phone.",
  "Really useful for tracking my steps and sleep.",
  "Would love more customization options.",
  "Absolutely love it — recommend to everyone.",
  "Too many steps to get anything done.",
  "Fine for what it is, nothing more.",
  "It helped me sleep better within a week."
)

categories <- c(
  "Positive: The respondent expresses satisfaction, approval, or favorable sentiment.",
  "Negative: The respondent expresses dissatisfaction, frustration, or criticism.",
  "Neutral: The respondent is factual, ambivalent, or does not express clear sentiment.",
  "Other: The response does not fit any of the above categories."
)

SURVEY_QUESTION <- "How do you feel about the new employee wellness program?"
```

## 1. Basic `classify()` without prompt tuning

Run a baseline classification first so you can compare accuracy before
and after tuning.

```{r baseline}
baseline <- classify(
  input_data      = responses,
  categories      = categories,
  research_question = SURVEY_QUESTION,
  api_key         = api_key,
  user_model      = "gpt-4o-mini"
)

cat("Baseline classifications:\n")
print(baseline[, c("input_data", grep("_label$", names(baseline), value = TRUE))])
```

## 2. `classify()` with `prompt_tune`

Pass `prompt_tune` (sample size) to enable automatic prompt optimization.
A browser window will open before `classify()` returns — correct any
misclassified rows and click **Submit**.

```{r prompt-tune}
# prompt_tune = 5   -- show 5 rows for correction per iteration
# tune_iterations = 1 -- one optimization pass (increase for more polish)
# tune_ui = 'browser' -- browser-based correction UI
# tune_optimize = 'balanced' -- maximize avg(accuracy, sensitivity, precision)

tuned <- classify(
  input_data        = responses,
  categories        = categories,
  research_question = SURVEY_QUESTION,
  api_key           = api_key,
  user_model        = "gpt-4o-mini",
  prompt_tune       = 5L,
  tune_iterations   = 1L,
  tune_ui           = "browser",
  tune_optimize     = "balanced"
)

cat("Tuned classifications:\n")
print(tuned[, c("input_data", grep("_label$", names(tuned), value = TRUE))])
```

When the browser window opens you will see a table with one row per
sampled response and one checkbox column per category. Check the boxes
that should be `1` for each response, then click **Submit**.

`prompt_tune` keeps the tuned prompt only if accuracy improves over the
baseline. If it does not improve, the original prompt is kept automatically
— no risk of making things worse.

## 3. Applying the tuned prompt to a full dataset

When you want to carry the optimized prompt to a separate `classify()` call
(e.g. to run on your full dataset after tuning on a subsample), call
`prompt_tune()` directly — it returns a list with a `system_prompt` field you
can pass to `classify()`.

```{r apply-tuned}
# Step A: tune on a small sample
tune_result <- prompt_tune(
  input_data      = responses,
  categories      = categories,
  api_key         = api_key,
  user_model      = "gpt-4o-mini",
  sample_size     = 5L,
  max_iterations  = 1L,
  ui              = "browser"
)

cat("Optimized system prompt:\n")
cat(tune_result$system_prompt, "\n")

# Step B: classify your full dataset with the tuned prompt
# (In practice, substitute full_df$response for `responses` here)
full_results <- classify(
  input_data        = responses,
  categories        = categories,
  research_question = SURVEY_QUESTION,
  api_key           = api_key,
  user_model        = "gpt-4o-mini",
  system_prompt     = tune_result$system_prompt
)

cat("Full-dataset classifications with tuned prompt:\n")
print(full_results[, c("input_data", grep("_label$", names(full_results), value = TRUE))])
```

## Tips

- **Start with `prompt_tune = 10`** and `tune_iterations = 2` for real
  datasets. A 5-row sample is fast but may miss edge cases.
- **Correct confidently.** Only fix rows you are certain about — uncertain
  corrections add noise that degrades the tuned prompt.
- **Use verbose category descriptions.** `prompt_tune` tunes the system
  prompt, not the category labels. Verbose labels (with "such as" clauses)
  give the meta-LLM more signal to work with.
- **Use a stronger tuning model.** The meta-LLM that writes the improved
  prompt benefits from higher capability — `gpt-4o` or `claude-opus` for
  tuning, even if you classify with `gpt-4o-mini`.
- **`tune_optimize = "sensitivity"`** is useful when false negatives are
  costly (e.g. safety-relevant categories). `"precision"` is useful when
  false positives are costly. `"balanced"` (default) optimizes the average
  of both.
- **Ollama works too.** Replace `user_model = "gpt-4o-mini"` with any
  local model and add `model_source = "ollama"`. Larger local models
  (qwen2.5:7b, mistral:7b) produce more useful corrections.
